---
title: "Example data analysis"
output: 
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Example data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, message=FALSE}
library(animation)
library(tidyverse)
library(here)
library(purrr)
# installed ffmpeg to show .gifs
```

### Question  

How has global temperature changed from 1880 to present? Are there temporal and geographic differences associated with temperature change? Can we model temperature increase linearly in recent years? 

```{r, fig.cap="The rate of Brownian motion is proportional to the temperature of a material. The warmer it is, the faster the particles move.", fig.show='animate', ffmpeg.format='gif', dev='jpeg', fig.width=6, fig.height=6}
ani.options(interval = 0.01, nmax = 150)
brownian.motion(pch = 21, 
                cex = 5, 
                col = "green", 
                bg = "#0072B2", 
                main = "Demonstration of Brownian Motion")
#saveGIF(brownian.motion(pch = 21, cex = 5, col = "darkgreen", bg = "#0072B2", main = "Demonstration of Brownian Motion"), movie.name="brownian_motion.gif")

```

### Data  

Data were discovered via [TidyTuesday (2023-07-11)](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-07-11). These data include global surface temperature observations from NASA from 1880 to 2023. The temperature anomalies (i.e. deviations from the average) are relative to 1951-1980 average temperatures. The data dictionary is available [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-07-11/readme.md#data-dictionary).



```{r, class.source = 'fold-show'}
# List required files to download from TidyTuesday
rds_files <- c("global_temps.rds", "nh_temps.rds","sh_temps.rds","zonann_temps.rds")

# Check if any of these files don't exist
if (any(!file.exists(here("data", rds_files)))) {
  # if data directory doesn't exist, make it
  if (!dir.exists(here("data"))) {
  dir.create(here("data")) }

# if files are missing, download the data
tuesdata <- tidytuesdayR::tt_load('2023-07-11')

global_temps <- tuesdata$global_temps
nh_temps <- tuesdata$nh_temps
sh_temps <- tuesdata$sh_temps
zonann_temps <- tuesdata$zonann_temps


# save the data object as a .csv file
saveRDS(global_temps, file = here("data","global_temps.rds"), version = 2)
saveRDS(nh_temps, file = here("data","nh_temps.rds"), version = 2)
saveRDS(sh_temps, file = here("data","sh_temps.rds"), version = 2)
saveRDS(zonann_temps, file = here("data","zonann_temps.rds"), version = 2)

}

# Load data
global_temps <- readRDS(here("data","global_temps.rds"))
nh_temps <- readRDS(here("data","nh_temps.rds"))
sh_temps <- readRDS(here("data","sh_temps.rds"))
zonann_temps <- readRDS(here("data","zonann_temps.rds"))

```


```{r}
# Wrangle data
zonann_temps_longer <- zonann_temps %>%
  pivot_longer(cols="Glob":"90S-64S",
               names_to="lat_band",
               values_to="ano")


```



```{r, fig.cap="Mean global temperature anomalies from 1880 to 2023, relative to 1951-1980 means. Temperature has increased in a non-linear way.", fig.width=7, fig.height=6}
# Plot data
zonann_temps_longer %>% 
  filter(lat_band=="Glob") %>%
  ggplot(aes(x = Year, y = ano)) +
  geom_line() +
  geom_smooth() +
  theme_minimal() +
  labs(y = "Global temperature anomaly (°C)",
       title = "Global temperature anomaly from 1880 to 2023",
       subtitle = "Global temperature has increased >1 °C since 1880",
       caption = "Data source: NASA GISS Surface Temperature Analysis (GISTEMP v4)")



```



```{r, fig.cap="Temperature anomalies from 1880 to 2023 in three latitudinal bands. Anomalies are relative to average temperatures from 1951-1980 (shaded in periwinkle). The overall magnitude and rate of temperature increase varies across latitudes.", fig.width=7, fig.height=6}
zonann_temps_longer %>%
  filter(lat_band %in% c("24N-90N","24S-24N","90S-24S")) %>%
  ggplot(aes(x = Year, y = ano)) +
  geom_point() +
  geom_line() +
  geom_rect(aes(xmin=1951, xmax=1980, ymin=-0.5, ymax=0.5), fill = "lightblue", alpha=0.01) +
  facet_wrap(~lat_band, nrow=3) +
  labs(y = "Temperature anomaly (°C)",
       title = "Temperature change by latitudinal band",
       subtitle = "Temperate and polar regions in the northern hemisphere experienced the largest swings \nand overall temperature change.",
       caption = "Data source: NASA GISS Surface Temperature Analysis (GISTEMP v4)") +
  theme_minimal()
  
# Calculate average anomaly for each column (i.e. different areas/latitude bands) in recent past (1970-2023)
mean_anos_by_lat_band <- zonann_temps %>%
  filter(Year>=1970) %>%
  select(-Year) %>%
  map_dbl(.f= mean)

```

Based on our plots, it appears that we have had a roughly linear increase in temperature from ~1970s to present. Let's use a linear model to estimate how fast temperature has increased over that period.

But first, let's review how a linear regression model is fit based on minimizing least squares loss function.


```{r, fig.cap="Demonstration of least squares minimization to fit the slope in linear regression.", fig.show='animate', ffmpeg.format='gif', dev='jpeg', fig.width=6, fig.height=6}
# Demo: least squares 
par(mar = c(5, 4, 2, 0.1))
ani.options(interval = 0.3, nmax = 50)

## default animation: with slope changing
least.squares(main = "Demo: Fit slope")

```

```{r, fig.cap="Demonstration of least squares minimization to fit the intercept in linear regression.", fig.show='animate', ffmpeg.format='gif', dev='jpeg', fig.width=6, fig.height=6}
# Demo: least squares 
par(mar = c(5, 4, 2, 0.1))
ani.options(interval = 0.3, nmax = 50)

## intercept changing
least.squares(ani.type = "intercept",
              main = "Demo: Fit intercept")

```


Ok, now let's fit our model.

```{r}
global_temp_df <- zonann_temps_longer %>%
  filter(lat_band=="Glob",
         Year>=1970)

temp.lm <- lm(ano ~ Year, data = global_temp_df)

#summary(temp.lm)
#plot(temp.lm)

# test quadratic model
# temp2.lm <- lm(ano ~ Year + I(Year*Year), data=global_temp_df)

global_temp_df$Predicted <- predict.lm(temp.lm, data = as.data.frame(seq(1970, 2023,1)))


```


```{r, fig.cap="Observed and predicted global temperature anomalies from 1970 to 2023, relative to 1951-1980 means. Temperature has increased in an approximately linear way over the past 50 years.", fig.width=7, fig.height=6}
global_temp_df_plotting <- global_temp_df %>%
  pivot_longer(ano:Predicted,
               names_to = "data_type",
               values_to = "value") %>%
  mutate(data_type = case_when(
    data_type == "ano" ~ "Observed",
    TRUE ~ data_type
  )) %>%
  select(-lat_band)
  
  
global_temp_df_plotting %>%
  ggplot(aes(x=Year, y = value, group = data_type, color = data_type)) +
  geom_line() +
  labs(y = "Temperature anomaly (°C)",
       title = "Global temperature change from 1970 to 2023",
       subtitle = "In recent years, global temperature has increased an average of ~0.02 °C per year.",
       caption = "Data source: NASA GISS Surface Temperature Analysis (GISTEMP v4)",
       color = "Data type") +
   theme_minimal()
  

```



### Summary  



### Functions used

dplyr: pivot_longer, filter, mutate, case_when, select
purrr: map_dbl, 
ggplot2: geom_smooth, geom_line, geom_point
animation: least.squares, brownian.motion

### TODO:
add purrr functions, 1 more dplyr/stringr functions, add explanatory text, add code-folding?

